Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-04-01 fscd19/response-fscd19.txt

If wished, we can remove the reference to the partial Agda
formalization from the paper.

- The choice of abstract syntax (Reviewer 3)

Reviewer 3 asks why we do not use the "familiar syntax [of] the
lambda-calculus instead of just the names of the rules?  Could
be some abstract syntax".

In fact we have given an abstract syntax since the rules correspond
1-1 to dependently typed constructors, thus a judgement with rules
corresponds 1-1 to a grammar for elements of an indexed set.  Disposing of
concrete syntax enables us to factor syntax conveniently, for instance
by using Cov in the definition of Nf (line 392).  In case of split₂
(line 438) it is not even immediate what the concrete-syntax analogy
of nesting should be.

@@

As reviewers 2 and 3 point out, the paper could have been presented
without any reference to category theory.  We use category theory just
as a source of terminology to accurately name the concepts we define.
Further, we like to include experts on category theory in our
readership, who could be upset about not naming the obvious.  The
paper is intended as a bridge between conventional approaches to
lambda-calculus and category-theoretic ones, without demanding too
much expertise.

- CBPV as logical system (Reviewer 2)

In the context of focused lambda-calculi, which we study to reduce the
degrees of liberty in normal forms (in other words, removing
don't-care non-determinism), CBPV arises naturally as calculus with
focusing on positive introductions but not on positive eliminations.
By "coincidence" (or rather not by coincidence, since there are no
coincidences in mathematics) it is the same calculus as the one found
in the search for a lambda-calculus with effects.

We are indeed only interested in CBPV as logical system.  Normalizing effects is
beyond the scope of this paper.  In any case, since effects do not follow the
introduction/elimination scheme, additional considerations would be necessary
for normalizing effects and would like likely depend on the individual effects
considered.

### 2019-04-01 fscd19/response-fscd19.txt

> where morally Γ ⊆ Δ gives an arrow Δ → Γ. This confused me more than
> once, especially at points where you said "this is a presheaf"
> omitting the category

In these techniques, one is always looking at a wide subcategory of the category of contexts and substitutions.
An order-preserving embedding Γ ⊆ Δ gives rise to a substitution Δ → Γ in the opposite direction, that's why we have to take presheaves over Cxt^op.
We agree that the direction of arrows in the category Cxt of contexts and order-preserving embeddings is non-standard from this perspective.
However, it has the benefit that our presheaves are just *covariant* functor J : Cxt → Set.

ADDED sentence explaining the relevance of Cxt^op.

@@

> 146: Maybe briefly say what a presheaf is to increase the
> accessibility of the paper even more?

FIXED. Made the introduction of presheaves more intuitive (thought still not
giving a general definition).

@@

FIXED. Also A should have been A ⇒ B.

>
> 217: I was confused by the 0 in the index for quite a while. This is a
> singleton context, isn't it? Then maybe write ε.0 or mention that
> singleton contexts look just like types

FIXED.

> 233: I feel repeating the category you build the monad on would
> help. It's on the category of presheaves over Cxt^{op}, right?
>
> 246: There are Γs missing twice on the reifications
>
> 253: the map should be indexed by Δ
>
> 261: again, the map needs a Δ
>
> 270: the Δ is the point your defining on pointwise, but that's too
> implicit currently in my opinion
>
> 315: Two sentences why products appear both negatively and positively
> might be nice for readers not familiar with polarized systems
>
> 319: "is not essential"

> [1] Rizkallah, Christine, Dmitri Garbuzov, and Steve Zdancewic.
> "A Formal Equational Theory for Call-By-Push-Value."
> International Conference on Interactive Theorem Proving. Springer, Cham, 2018.
>
> [2] Forster, Yannick, et al.
> "Call-by-push-value in Coq: operational, equational, and denotational theory."
> Proceedings of the 8th ACM SIGPLAN International Conference on Certified Programs and Proofs. ACM, 2019.
>
> ----------------------- REVIEW 3 ---------------------
> PAPER: 50
> TITLE: Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus
> AUTHORS: Andreas Abel and Christian Sattler
>
> Overall evaluation: 1 (weak accept)
>
> ----------- Overall evaluation -----------
>
> Normalization by Evaluation (NbE) is the (by now well-understood)
> beautiful idea that we can normalize terms in a typed lambda-calculus
> through mutually recursive functions for evaluation and reification,
> partially based on types.  NbE is one way to relate programming
> languages (where we don't evaluate under lambda-abstractions) and pure
> proof theory (where we fully normalize).
>
> Sums however present a difficulty.  In Section 2 the present paper
> adapts a prior idea on using monads to update the semantic domain
> needed for NbE to capture sums.  I find this idea compelling and
> NbE factors very cleanly for a lambda-calculus with products and
> sums.
>
> This, however, is characterized as "wasteful" (although I did not
> understand what the authors meant by this), so Section 3 polarizes
> the types and restructures NbE so that the use of the monad is
> confined to the transition from positive to negative types.  This
> seems like a natural idea, clarifying the approach of Section 2.
> The simplicity of the definitions at the end of this section is
> once again compelling, although some complexity is hidden in the
> definition of the semantic functions.
>
> Again, the authors express some dissatisfaction I couldn't quite
> follow and move on to a polarized and focused presentation.  This
> step makes perfect sense to me, although it was difficult for me
> to extract additional insight from this section.
>
> I am not familiar enough with the prior work to judge the degree
> of originality confidently, but the idea of using a monad to
> handle sums in NbE in this manner was new to me, as was the
> extension to the polarized language.  The results have been
> (mostly?) formalized, so I don't have any reason to doubt their
> correctness.  NbE is certainly interesting and I find the results
> to be significant in its further study, even if the open problems
> mentioned in the conclusion are left to future work.
>
> The presentation is idiosyncratic in multiple ways, which one may
> view as either charming or annoying.  Ultimately, I care if ideas
> are communicated well so they can be easily understood, and for me
> the paper receives mixed grades on that score.  I am going into
> some detail on issues of the presentation below.  Overall, I like
> the subject, the paper itself, and I believe I understood the main
> ideas, but I think the presentation can be improved to make it
> more understandable with less background.
>
> Presentation issues:
>
> Intrinsic typing.  Certainly an acceptable choice and seems to
> work well in the formalization.
>
> No syntax.  That presented a significant obstacle to me.  Why not
> use familiar syntax the the lambda-calculus instead of just the
> names of the rules?  Could be some abstract syntax, but even that
> would have helped.

### 2019-04-01 fscd19/fscd19.tex
Minor changes.

By a presheaf, we will always mean a presheaf over the category $\Cxt\op$, i.e.,
a functor from $\Cxt$ to $\Set$.  For example, given any type $A$, we can
consider $A{\in}\_$ as such a presheaf, with action on objects mapping $\Gamma$
to the set $A \in \Gamma$ of indices of $A$ in $\Gamma$ and action on morphisms
witnessed by $\treindex : \Gamma \sublist \Delta \to A \in \Gamma \to A \in
\Delta$.  The associated functor laws $\treindex\,\tid\,x \equiv x$ and
$\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv \treindex\,(\tau_1 \rcomp
\tau_2)\,x$ hold propositionally.

@@

CBPV calls positive types $P$ \emph{value types} $A$ and negative
types $N$
\emph{computation types} $\underline B$,
yet we shall stick to our terminology
which is common in publications on focalization.
However, we shall use $\tThunk$ for switch $\downarrow$ and $\tComp$ for
switch $\uparrow$.
\[
\begin{array}{lllrl@{\qquad}l}
  \PTy & \ni & P,Q & ::= & \patom
    \mid \tyone \mid P_1 \times P_2
    \mid \tyzero \mid P_1 + P_2
    \mid \Thunk N
  & \mbox{positive type}
\\
  \NTy & \ni & N,M & ::= & \natom
    \mid \top \mid N_1 \ntimes N_2
    \mid P \To N
    \mid \Comp P
  & \mbox{negative type}
\\
\end{array}
\]
CBPV uses $U$ for $\tThunk$ and $F$ for $\tComp$, however, we find
these names uninspiring unless you have good knowledge of the intended
model.  Further, CBPV employs labeled sums $\Sigma_I (P_i)_{i : I}$
and labeled records $\Pi_I (P_i)_{i : I}$ for up to countably infinite
label sets $I$ while we only have finite sums $(\tyzero,+)$ and
records $(\top,\ntimes)$.  However, this difference in not essential,
our treatment extends directly to the infinite case, since we are
working in type theory which allows infinitely branching inductive
types.   As a last difference, CBPV does not consider base types; in
anticipation of the next section, we add them as both positive atoms
($\patom$) and negative atoms ($\natom$).

>>>>

By a \emph{presheaf} we will always mean a presheaf over the category $\Cxt\op$,
i.e., a functor from $\Cxt$ to $\Set$.\footnote{With respect to the terms
defined later, $\Cxt\op$ is to be seen as a category of contexts and those
substitutions given by weakening.}  For example, given any type $A$, we can
consider $A{\in}\_$ as such a presheaf, with action on objects mapping $\Gamma$
to the set $A \in \Gamma$ of indices of $A$ in $\Gamma$ and action on morphisms
witnessed by $\treindex : \Gamma \sublist \Delta \to A \in \Gamma \to A \in
\Delta$.  The associated functor laws $\treindex\,\tid\,x \equiv x$ and
$\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv \treindex\,(\tau_1 \rcomp
\tau_2)\,x$ hold propositionally.

@@

CBPV calls positive types $P$ \emph{value types} $A$ and negative
types $N$
\emph{computation types} $\underline B$,
yet we shall stick to our terminology
which is common in publications on focalization.
However, we shall use $\tThunk$ for switch $\downarrow$ and $\tComp$ for
switch $\uparrow$.
\[
\begin{array}{lllrl@{\qquad}l}
  \PTy & \ni & P,Q & ::= & \patom
    \mid \tyone \mid P_1 \times P_2
    \mid \tyzero \mid P_1 + P_2
    \mid \Thunk N
  & \mbox{positive type}
\\
  \NTy & \ni & N,M & ::= & \natom
    \mid \top \mid N_1 \ntimes N_2
    \mid P \To N
    \mid \Comp P
  & \mbox{negative type}
\\
\end{array}
\]
CBPV uses $U$ for $\tThunk$ and $F$ for $\tComp$, however, we find
these names uninspiring unless you have good knowledge of the intended
model.  Further, CBPV employs labeled sums $\Sigma_I (P_i)_{i : I}$
and labeled records $\Pi_I (P_i)_{i : I}$ for up to countably infinite
label sets $I$ while we only have finite sums $(\tyzero,+)$ and
records $(\top,\ntimes)$.  However, this difference is not essential,
our treatment extends directly to the infinite case, since we are
working in type theory which allows infinitely branching inductive
types.   As a last difference, CBPV does not consider base types; in
anticipation of the next section, we add them as both positive atoms
($\patom$) and negative atoms ($\natom$).
